---
title: "Evaluation of surgery time and short-term incision healing for Largemouth Bass using smooth and barbed absorbable monofilament suture types."
author: "Jeff Stevens"
date: "2024-05-02"
output: html_document
---


### Load Packages ###

Lets start by loading the packages that we will use to analyze and visualize our data. Note that all messages are "turned off" for this code chunk. 

```{r, echo = FALSE, warning = FALSE, message = FALSE}

require(readr) # this package will help us download our data from GitHub
require(tidyverse) # this package will help us visualize our data
require(ggpubr) # this package will help us save our figures in a publishable format
require(car) # this package will help our regressions
library(lmtest) # this package will help us with likelihood ratio tests

```

### Load Data ###

Next, we will load our surgical data from GitHub. One fish (fish 0814) was released due to collection stress, so we will exclude it from our analyses.

```{r}

# Load data from GitHub

datafull = read_csv("https://raw.githubusercontent.com/StevensFishEco/SutureComparison/main/Suture_Surgery_Data.csv", show_col_types = FALSE)

# Remove fish 0814

datafull <- datafull[-14,]

```

### Data Visualization ###

#### ggplot2 Theme ####

Before we analyze our data, lets visualize it. First, we will define an overall theme that we will use for all plots. This theme produces AFS styled figures in ggplot2 (Glassic et al. 2019). The original theme code is available at https://github.com/CGuyMSU/AFS-figures.

```{r, echo = FALSE, warning = FALSE, message = FALSE}

# AFS styled ggplot2 theme from Glassic et al. (2019).

theme_AFS <- function() {
  theme_classic() + 
  theme(axis.title.y = element_text(size = 14, margin = margin(t = 0, r = 10, b = 0, l = 0), colour = "black"), axis.title.x = element_text(size = 14, margin = margin(t = 10, r = 0, b = 0, l = 0), colour = "black"), text = element_text(family = "Times"), plot.title = element_text(face = "bold", family = "Arial"), legend.position = c(0.3,0.85), legend.text = element_text(size = 12), plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"), axis.ticks.y = element_line(size = 0.5), axis.ticks.x = element_line(size = 0.5), axis.ticks.length = unit(0.2,"cm"), axis.text.y = element_text(colour = "black", size = 12, angle = 0, vjust = 0.5, hjust = 1, margin = margin(t = 0, r = 5, b = 0, l = 0)), axis.text.x = element_text(colour = "black", size = 12, angle = 0, vjust = 0, hjust = 0.5, margin = margin(t = 5, r = 0, b = 0, l = 0)), axis.line = element_line(colour = "black", size = 0.5, lineend = "square"))}

```

#### Surgery and Recovery Time ####

We will start by looking at our surgery and recovery times (recorded in seconds). We will use two different plots to visualize each. First, we will visualize surgery and recovery time by surgeon (i.e., surgeon A, B, or C). Second, we will visualize surgery and recovery time by suture material (i.e., barbed and smooth sutures). We will do this with simple boxplots (in the style of Tukey).

```{r}

#### Surgery Times ####

# By surgeon

a1 <- ggplot(data = datafull) +
  aes(x = Surgeon, y = Surgery_Secs) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +
  theme_AFS() +
  theme(legend.position = "none", axis.text.x = element_blank(), axis.title.x = element_blank(), plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm")) +
  labs(fill = "Suture type") +
  scale_y_continuous(breaks = seq(0, 480, 240), lim = c(0, 480)) +
  xlab("") +
  ylab("Surgery time")

# By suture type

b1 <- ggplot(data = datafull) +
  aes(x = Suture_Type, y = Surgery_Secs) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +
  theme_AFS() +
  theme(legend.position = "none", axis.text.x = element_blank(), axis.text.y = element_blank(), axis.title.x = element_blank(), axis.title.y = element_blank(), plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm")) +
  scale_y_continuous(breaks = seq(0, 480, 240), lim = c(0, 480)) +
  xlab("") +
  ylab("")

#### Recovery Times ####

# By surgeon

c1 <- ggplot(data = datafull) +
  aes(x = Surgeon, y = Recovery_Secs) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +
  theme_AFS() +
  theme(legend.position = "none", plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm"), axis.title.y = element_text(size = 14, margin = margin(t = 0, r = 14, b = 0, l = 0), colour = "black")) +
  scale_y_continuous(breaks = seq(0, 60, 30), lim = c(0, 60)) +
  xlab("Surgeon") +
  ylab("Recovery time")

# By suture type

d1 <- ggplot(data = datafull) +
  aes(x = Suture_Type, y = Recovery_Secs) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +
  theme_AFS() +
  theme(legend.position = "none", axis.text.y = element_blank(), axis.title.y = element_blank(), plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm")) +
  scale_y_continuous(breaks = seq(0, 60, 30), lim = c(0, 60)) +
  xlab("Suture type") +
  ylab("")

#### Combine plots ####

fig1 <- ggarrange(ggarrange(a1, b1, labels = c("A","B"), ncol = 2, nrow = 1, align = "h", widths = c(1,0.85), font.label = list(size = 12, color = "black", face = "plain", family = "Times")), ggarrange(c1, d1, labels = c("C","D"), ncol = 2, nrow = 1, align = "h", widths = c(1,0.85), font.label = list(size = 12, color = "black", face = "plain", family = "Times")), nrow = 2, heights = c(0.85, 1), align = "none")

fig1

# We'll save it as a .tiff file.

#ggsave(filename = "Figure01_Boxplot.tiff", plot = fig1, scale = 1, width = 7.25, height = 4.8, units = "in", dpi = "retina")

```

#### Openness and Inflammation ####

Now, we will look at the openness and inflammation scores. As above, we will use two different plots to visualize each. First, we will visualize openness and inflammation by surgeon (i.e., surgeon A, B, or C). Second, we will visualize openness and inflammation by suture material (i.e., barbed and smooth sutures).

```{r}

#### Openness ####

# By surgeon

a2 <- ggplot(data = datafull) +
  aes(x = Surgeon, y = Open) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +  
  theme_AFS() +
  theme(legend.position = "none", axis.text.x = element_blank(), axis.title.x = element_blank(), plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm")) +
  labs(fill = "Suture type") +
  scale_y_continuous(breaks = seq(0, 4, 2), lim = c(0, 4)) +
  xlab("") +
  ylab("Openness")

# By suture type

b2 <- ggplot(data = datafull) +
  aes(x = Suture_Type, y = Open) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +  
  theme_AFS() +
  theme(legend.position = "none", axis.text.x = element_blank(), axis.text.y = element_blank(), axis.title.x = element_blank(), axis.title.y = element_blank(), plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm")) +
  scale_y_continuous(breaks = seq(0, 4, 2), lim = c(0, 4)) +
  xlab("") +
  ylab("")

#### Inflammation ####

# By surgeon

c2 <- ggplot(data = datafull) +
  aes(x = Surgeon, y = Inflame) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +
  theme_AFS() +
  theme(legend.position = "none", plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm"), axis.title.y = element_text(size = 14, margin = margin(t = 0, r = 14, b = 0, l = 0), colour = "black")) +
  scale_y_continuous(breaks = seq(0, 4, 2), lim = c(0, 4)) +
  xlab("Surgeon") +
  ylab("Inflammation")

# By suture type

d2 <- ggplot(data = datafull) +
  aes(x = Suture_Type, y = Inflame) +
  geom_boxplot(width = 0.8, outlier.shape = 21) +
  theme_AFS() +
  theme(legend.position = "none", axis.text.y = element_blank(), axis.title.y = element_blank(), plot.margin = unit(c(0.1, 0.1, 0.5, 0.6), "cm")) +
  scale_y_continuous(breaks = seq(0, 4, 2), lim = c(0, 4)) +
  xlab("Suture type") +
  ylab("")

#### Combine plots ####

fig2 <- ggarrange(ggarrange(a2, b2, labels = c("A","B"), ncol = 2, nrow = 1, align = "h", widths = c(1,0.85), font.label = list(size = 12, color = "black", face = "plain", family = "Times")), ggarrange(c2, d2, labels = c("C","D"), ncol = 2, nrow = 1, align = "h", widths = c(1,0.85), font.label = list(size = 12, color = "black", face = "plain", family = "Times")), nrow = 2, heights = c(0.85, 1), align = "none")

fig2

# We'll save it as a .tiff file.

#ggsave(filename = "Figure02_Boxplot.tiff", plot = fig2, scale = 1, width = 7.25, height = 4.8, units = "in", dpi = "retina")

```

#### Recovery Time by Surgery Time ####

Since recovery time should be related to surgery time, we will visualize this relationship with a scatterplot and use a simple linear model to create a regression line in the plot.

```{r}

g <- ggplot(data = datafull) +
  aes(x = Surgery_Secs, y = Recovery_Secs) +
  geom_point(size = 2, aes(shape = Suture_Type)) +
  geom_smooth(method = "lm", color = "black") +
  theme_AFS() +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(90, 450, 180), lim = c(90, 450)) +
  scale_y_continuous(breaks = seq(0, 60, 30), lim = c(0, 60)) +
  xlab("Surgery Time (seconds)") +
  ylab("Recovery Time (seconds)")

g

```

### Statistical Analyses ###

Now that we have visualized our data, lets test if any of the trends we saw are statistically significant.

#### Initial Length ####

First, we will test if there was any differences in initial length between treatment groups.

```{r}

median(datafull$Tl_mm)
min(datafull$Tl_mm)
max(datafull$Tl_mm)

t.test(Tl_mm ~ Suture_Type, data = datafull) # by suture type
summary(aov(Tl_mm ~ Surgeon, data = datafull)) # by surgeon

```

#### Surgery Time ####

Next, we will test if there was any differences in surgery time between suture types or among surgeons.

```{r}

median(datafull$Surgery_Secs)
min(datafull$Surgery_Secs)
max(datafull$Surgery_Secs)

mod10 <- aov(Surgery_Secs ~ Surgeon + Suture_Type + Surgeon*Suture_Type, data = datafull) 
summary(mod10) # by suture type and surgeon with interaction

mod11 <- aov(Surgery_Secs ~ Surgeon + Suture_Type, data = datafull) 
summary(mod11) # by suture type and surgeon without interaction

```

#### Recovery Time ####

Next, we will test if there was any differences in recovery time between suture types or among surgeons. Then, we will test if recovery time was related to surgery time.

```{r}

median(datafull$Recovery_Secs)
min(datafull$Recovery_Secs)
max(datafull$Recovery_Secs)

mod20 <- aov(Recovery_Secs ~ Surgeon + Suture_Type + Surgeon*Suture_Type, data = datafull)
summary(mod20) # by suture type and surgeon with interaction

mod21 <- aov(Recovery_Secs ~ Surgeon + Suture_Type, data = datafull)
summary(mod21) # by suture type and surgeon without interaction

mod22 <- glm(Recovery_Secs ~ Surgery_Secs, family = Gamma, data = datafull)
summary(mod22) # by surgical time
lrtest(mod22)

```

#### Mortality ####

Next, we will test if there was any differences in mortality between suture types or among surgeons. Then, we will test if mortality was related to surgery time or recovery time.

```{r}

mod30 <- glm(Mortality ~ Suture_Type, family = binomial, data = datafull)
summary(mod30) # by suture type
lrtest(mod30)

mod31 <- glm(Mortality ~ Surgeon, family = binomial, data = datafull)
summary(mod31) # by surgeon
lrtest(mod31)

mod32 <- glm(Mortality ~ Surgery_Secs, family = binomial, data = datafull)
summary(mod32) # by surgical time
lrtest(mod32)

mod33 <- glm(Mortality ~ Recovery_Secs, family = binomial, data = datafull)
summary(mod33) # by recovery time
lrtest(mod33)

```

#### Openness ####

Next, we will test if there was any differences in openness between suture types or among surgeons. Then, we will test if openness was related to surgery time or recovery time.

```{r}

median(na.omit(datafull$Open))
min(na.omit(datafull$Open))
max(na.omit(datafull$Open))

mod40 <- aov(Open ~ Surgeon + Suture_Type + Surgeon*Suture_Type, data = datafull)
summary(mod40) # by suture type and surgeon with interaction

mod41 <- aov(Open ~ Surgeon + Suture_Type, data = datafull)
summary(mod41) # by suture type and surgeon without interaction

mod42 <- glm(Open ~ Surgery_Secs, data = datafull)
summary(mod42) # by surgical time
lrtest(mod42)

mod43 <- glm(Open ~ Recovery_Secs, data = datafull)
summary(mod43) # by recovery time
lrtest(mod43)

```

#### Inflammation ####

Finally, we will test if there was any differences in inflammation between suture types or among surgeons. Then, we will test if inflammation was related to surgery time or recovery time.

```{r}

median(na.omit(datafull$Inflame))
min(na.omit(datafull$Inflame))
max(na.omit(datafull$Inflame))

mod50 <- aov(Inflame ~ Surgeon + Suture_Type + Surgeon*Suture_Type, data = datafull)
summary(mod50) # by suture type and surgeon with interaction

mod51 <- aov(Inflame ~ Surgeon + Suture_Type, data = datafull)
summary(mod51) # by suture type and surgeon without interaction

DunnettTest(x = datafull$Inflame, g = datafull$Surgeon)

mod52 <- glm(Inflame ~ Surgery_Secs, data = datafull)
summary(mod52) # by surgical time
lrtest(mod52)

mod53 <- glm(Inflame ~ Recovery_Secs, data = datafull)
summary(mod53) # by recovery time
lrtest(mod53)

```
